Beyond the Numbers: Dissecting New York City's Leading Causes of Death Across Demographics

### 2024-03-17 paper/paper.qmd
updated models and paper

The causes of mortality and determinants of such mortality causes have puzzled researchers for centuries. Despite evidence of the cause of mortality identification inaccuracies (Johns et al., 2013), comprehensive and detailed death certificate records along with more powerful computing and statistical analysis tools have made it much easier to observe actual causes of mortality – at least in the developed world. The determinants of mortality are much harder to pinpoint, with the relative strength of each determinant an even more daunting task.
A popular theory posits that “the wealthier the healthier” – that is, higher incomes directly cause lower mortality, and that higher incomes are the leading cause of lower mortality. Yet this supposed causal relationship can likewise work in reverse in which higher mortality means lower incomes, and studies have shown this explains much of the correlation (Cutler et al., 2006, p. 117). Surprisingly it has also been observed that recessions seem to improve health outcomes, as people are more likely to do less health-damaging activities (i.e., drinking and smoking) and more health-enhancing activities (i.e., exercise) (Ruhm, 2000). Another theory suggests that education and ethnicity are the leading causes of lower mortality. For example, according to a study, higher mortality is observed among children of black mothers relative to children of white mothers, regardless of the education level of the mother. However, when comparing only children of white mothers, mothers with higher education levels are observed to show much lower relative mortality for their children (Pamuk et al., 1998). Higher education levels among women are consistently correlated with lower mortality from cardiovascular disease and lung cancer, but also higher mortality from breast cancer (Cutler et al., 2006, p. 113) – demonstrating the importance of comparing mortality by cause. From a long-run perspective, there was hardly any observed difference in life expectancy among England's rich and poor before the Enlightenment period. Yet past this period, there is a widespread general divergence. Since education has always been more accessible for the rich post-Enlightenment, this highlights the strong potential causal link between education (i.e., the development of new ideas of personal and public health such as germ theory of disease, and openness to adjust behaviours or implement policies validating such ideas) and mortality (Cutler et al., 2006, p. 117). Finally, there is a theory that explains differences in mortality by cause due to differing socio-cultural practices by ethnicity and education levels. For example, lung cancer and cardiovascular diseases should be lower among highly educated people since they are less likely to smoke (Pamuk et al., 1998). While this theory has its merits, the size of its effect is questionable. For example, smoking prevalence is higher among white Americans relative to black Americans, yet this cannot explain black-white differences in American mortality (Cutler et al., 2006, p. 114).

>>>>

The causes of mortality and determinants of such mortality causes have puzzled researchers for centuries. Despite evidence of the cause of mortality identification inaccuracies {Johns2013}, comprehensive and detailed death certificate records along with more powerful computing and statistical analysis tools have made it much easier to observe actual causes of mortality – at least in the developed world. The determinants of mortality are much harder to pinpoint, with the relative strength of each determinant an even more daunting task.
A popular theory posits that “the wealthier the healthier” – that is, higher incomes directly cause lower mortality, and that higher incomes are the leading cause of lower mortality. Yet this supposed causal relationship can likewise work in reverse in which higher mortality means lower incomes, and studies have shown this explains much of the correlation {Cutler2006}. Surprisingly it has also been observed that recessions seem to improve health outcomes, as people are more likely to do less health-damaging activities (i.e., drinking and smoking) and more health-enhancing activities (i.e., exercise) {Ruhm, 2000}. Another theory suggests that education and ethnicity are the leading causes of lower mortality. For example, according to a study, higher mortality is observed among children of black mothers relative to children of white mothers, regardless of the education level of the mother. However, when comparing only children of white mothers, mothers with higher education levels are observed to show much lower relative mortality for their children {Pamuk1998}. Higher education levels among women are consistently correlated with lower mortality from cardiovascular disease and lung cancer, but also higher mortality from breast cancer {Cutler2006} – demonstrating the importance of comparing mortality by cause. From a long-run perspective, there was hardly any observed difference in life expectancy among England's rich and poor before the Enlightenment period. Yet past this period, there is a widespread general divergence. Since education has always been more accessible for the rich post-Enlightenment, this highlights the strong potential causal link between education (i.e., the development of new ideas of personal and public health such as germ theory of disease, and openness to adjust behaviours or implement policies validating such ideas) and mortality {Cutler2006}. Finally, there is a theory that explains differences in mortality by cause due to differing socio-cultural practices by ethnicity and education levels. For example, lung cancer and cardiovascular diseases should be lower among highly educated people since they are less likely to smoke {Pamuk1998}. While this theory has its merits, the size of its effect is questionable. For example, smoking prevalence is higher among white Americans relative to black Americans, yet this cannot explain black-white differences in American mortality {Cutler2006}.
